Applying Step Selection Functions to Data with Data Gaps: Simulation and Case Study on African wild dogs

### 2022-07-20 Chapter_4/04_Manuscript/Manuscript.tex
updated scripts

\textbf{Keywords:} step selection functions, gps data, animal movement, Lycaon
pictus, irregular data

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
This is an abstract
\end{abstract}

@@

\section{Introduction}
GPS data is usually collected on fixed temporal intervals (could do a quick and
dirty analysis on MoveBank) ranging from a few minutes, to multiple hours or
even days between subsequent reolocations. Nevertheless, GPS regularly fail to
adhere to the anticipated fixrate schedule due to cloud cover or other
unfavorable conditions, thus resulting in irregular data. In other cases, the
GPS schedule is not strictly fixed but may be subject to change. For example,
\cite{Cozzi.2020} increase the fixrate of their GPS collars once per week from 4
to 8 fixes per day. We chose to coin this type of data irregular, because it is
not collected on the anticipated fixrate schedule and therefore not directly
comparable to the vast majority of collected data.

Step selection analysis is a powerful framework that enables researchers to
study habitat and movement preferences of their focal species. The method works
by comparing spatial covariates at locations where the studied animal has been
observed, to a set of locations where the animal was presumably absent. For
this, the collected GPS data is converted into steps, where a step represents
the straight line movement between two consecutive GPS fixes. One can then
compare covariates experienced along observed steps to covariates along
potential alternative steps and thus infer preferred or avoided features. By
restricting the availability domain to a set of steps, SSA is specifically
intended to account for spatio-temporal autocorrelation inherent to data
collected using GPS collars and is therefore one of the preferred methods for
analysing such data. However, it is considered good practice to only consider
steps with similar step durations in step-selection models. Depending on the
amount of available data and depending on the number of missing GPS fixes, this
might entail little sacrifice. In some situations, however, removing irregular
data implies a substantial loss of information. Additionally, recent
developments in the SSA realm have brought forward several improvements to SSA
that might allow to relax the assumption of regular step-durations. This
includes the \textit{integrated} SSA approach presented by \citep{Avgar.2016},
in which inference on habitat \textit{and} movement preferences are possible
thanks to the inclusion of movement descriptors in the respective model. More
recently, the method has been further refined by \citep{Munden.2021}, who coined
the term \textit{time-varying integrated} SSA. This method was developed for
high frequency data which has been rarified using a change-point detection
algorithm, thus also resulting in irregular step-durations. In case of missing
steps, in contrast, steps durations are not entirely random but usually
clustered.

@@

@@

\section{Results}
\section{Discussion}
Results from our simulation study demonstrate that the inclusion of GPS fixes
can be used to gain information on relative habitat preferences of the focal
species. While the inclusion of steps from irregular sampling schemes resulted
in substantial biases for estimates related to step metrics, we showed that
these biases can effectively be mitigated by generating step lengths and turning
angles from distributions that are separately fit to steps from different step
durations. Regardless of biases in beta coefficients for step metrics, we found
that habitat selection estimates were unbiased, even when including irregular
steps. Even more, the inclusion of irregular steps drastially reduced model
uncertainty, thus suggesting that such data can be used to gain information that
would otherwise be lost.

>>>>

\textbf{Keywords:} step selection analysis, gps data, animal movement, irregular
data

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Animal movement data collected using GPS collars is often noisy and incomplete,
thus confronting movement ecologists with data that is not always uniformly
distributed in time. Step selection analysis has become a popular tool to
analyse such data and to infer a species movement and habitat preferences, yet
it requires that consecutive GPS relocations are collected at regular temporal
intervals. To adhere to this requirement, researchers typically discard large
portions of their data and only consider GPS bursts that were collected at
comparable intervals. Here, we assess the appropriateness of this practice and
use simulated data to examine if including temporally irregular data could, in
fact, improve estimates from step-selection analyses. Moreover, we explore the
usefulness of several methods to account for the temporal irregularity. Our
results suggest that utilizing a full dataset, despite potential temporal
irregularities, is beneficial, especially when an appropriate method is chosen
to account for such irregularities.
\end{abstract}

@@

\section{Introduction}
GPS data is usually collected on fixed temporal intervals (could do a quick and
dirty analysis on MoveBank) ranging from a few minutes, to multiple hours or
even days between subsequent reolocations. Nevertheless, GPS devices frequently
fail to adhere to the anticipated schedule, for instance due to cloud cover or
other unfavorable conditions, thus resulting in irregular data. In other cases,
the GPS schedule is not strictly fixed but may be subject to change. For
example, \cite{Cozzi.2020} increased the fixrate of their GPS collars once per
week from 4 to 8 fixes per day, aiming to collect more fine scaled data for a
limited time. We chose to coin this type of data \textit{irregular}, because it
is not collected on the anticipated fixrate schedule and therefore not directly
comparable to the vast majority of the collected data.

Step selection analysis is a powerful framework that enables researchers to
study habitat and movement preferences of their focal species. The method works
by comparing spatial covariates at locations where the studied animal has been
observed, to a set of locations where the animal was presumably absent. For
this, the collected GPS data is converted into steps, where a step represents
the straight line movement between two consecutive GPS fixes. One can then
compare covariates experienced along observed steps to covariates along
potential alternative steps and thus infer preferred or avoided features. By
restricting the availability domain to a set of steps, SSA is specifically
intended to account for spatio-temporal autocorrelation inherent to data
collected using GPS collars and is therefore one of the preferred methods for
analysing such data. However, it is considered good practice to only consider
steps with similar step durations in step-selection models. Depending on the
amount of available data and depending on the number of missing GPS fixes, this
practice might entail little sacrifice. In some situations, however, removing
irregular data implies a substantial loss of information. Additionally, recent
developments in the SSA realm have brought forward several improvements to SSA
that might allow to relax the assumption of regular step-durations. This
includes the \textit{integrated} SSA approach presented by \citep{Avgar.2016},
in which inference on habitat \textit{and} movement preferences are possible
thanks to the inclusion of movement descriptors in the respective model. More
recently, the method has been further refined by \citep{Munden.2021}, who coined
the term \textit{time-varying integrated} SSA. This method was developed for
high frequency data which has been rarified using a change-point detection
algorithm, thus also resulting in irregular step-durations. In case of missing
steps, in contrast, steps durations are not entirely random but usually
clustered.

@@

Specifically, we employed six competing approaches to handle temporal
irregularity in the data.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_NumberOfSteps.png}
  \caption{}
  \label{Covariates}
  \end{center}
\end{figure}

@@

\section{Results}

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_DistributionParameters.png}
  \caption{}
  \label{DistributionParameters}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_ResultsHabitatKernel.png}
  \caption{}
  \label{ResultsHabitat}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_ResultsMovementKernel.png}
  \caption{}
  \label{ResultsMovement}
  \end{center}
\end{figure}

\section{Discussion}
Results from our simulation study demonstrate that the inclusion of GPS fixes
can be used to gain information on relative habitat preferences of the focal
species. While the inclusion of steps from irregular sampling schemes resulted
in substantial biases for estimates related to step metrics, we showed that
these biases can effectively be mitigated by generating step lengths and turning
angles from distributions that are separately fit to steps from different step
durations. Regardless of biases in beta coefficients for step metrics, we found
that habitat selection estimates were unbiased, even when including irregular
steps. Even more, the inclusion of irregular steps drastially reduced model
uncertainty, thus suggesting that such data can be used to gain information that
would otherwise be lost.
